# Reproducible Research: Peer Assessment 1

## Loading and preprocessing the data
### The script is executed on the same directory of activity.zip is stored
```{r cache=TRUE, echo=TRUE}
unzip(zipfile="activity.zip")

activity  <- read.csv(file="activity.csv")
dim(activity)
head(activity)
summary(activity)
```

### Creates a new dataframe excluding NA values for the variable "steps"
```{r echo=TRUE}
activity.notNA  <- activity[!is.na(activity$steps),]
dim(activity.notNA)
summary(activity.notNA)
```

## What is mean total number of steps taken per day?
### Totalize the number of steps per day
```{r echo=TRUE}
require(data.table)
steps.dt  <- data.table(activity.notNA, key="date")
total.steps  <- steps.dt[,list(steps=sum(steps)), by=date]
```

### Make a histogram of the total number of steps taken each day
```{r echo=TRUE}
hist(total.steps$steps, main="Total number of steps taken each day", xlab="Steps")
```

### Calculates and prints mean and median of the total number of steps taken per day
```{r echo=TRUE}
print(paste("Steps. Mean:", mean(total.steps$steps)))
print(paste("Steps. Median:", median(total.steps$steps)))
```

## What is the average daily activity pattern?
### Averages the number of steps per 5-minute interval
```{r echo=TRUE}
steps.dt.av  <- data.table(activity.notNA, key="interval")
average.steps  <- steps.dt.av[,list(steps=mean(steps)), by=interval]
```

### Make a time series plot of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)
```{r echo=TRUE}
plot(x=average.steps$interval, y=average.steps$steps, type="l", 
     , xlab="Steps. Average", ylab="Interval", main="Average number of steps taken each 5-minute interval")
```

## Imputing missing values

### Calculates and reports the total number of missing values in the dataset
```{r}
print(paste("Number of NAs in the dataset:", sum(is.na(activity))))
```

### Filling in all of the missing values in the dataset with the mean for that 5-minute interval
```{r}
# Saving index of NAs
nas.index  <- activity[which(is.na(activity$steps)), ]

# Looping over NAs and getting the mean value
steps.meanValues  <- by(activity[is.na(activity$steps), ],  1:nrow(nas.index), simplify=T
   , function(row) average.steps[average.steps$interval==row$interval,steps])
```

### Create a new dataset that is equal to the original dataset but with the missing data filled in.
```{r}
# Cloning original dataframe
activityNew  <- activity

# Setting NA steps with the mean
activityNew[which(is.na(activityNew$steps)), 1]  <-  steps.meanValues 

print(paste("Number of NAs in the new dataset:", sum(is.na(activityNew))))
```

### Make a histogram of the total number of steps taken each day 
```{r}
steps.dt.new  <- data.table(activityNew, key="date")
total.steps.new  <- steps.dt.new[,list(steps=sum(steps)), by=date]
hist(total.steps.new$steps, main="Total number of steps taken each day", xlab="Steps")
```

### Calculate and report the mean and median total number of steps taken per day.
```{r}
print(paste("Steps. Mean:", mean(total.steps.new$steps)))
print(paste("Steps. Median:", median(total.steps.new$steps)))
```

### Do these values differ from the estimates from the first part of the assignment? 
```{r}
print(paste("Difference in the mean:", mean(total.steps.new$steps)-mean(total.steps$steps)))
print(paste("Difference in the median:", median(total.steps.new$steps)-median(total.steps$steps)))
```

### What is the impact of imputing missing data on the estimates of the total daily number of steps?
```{r}
sum.difference  <- sum(total.steps.new$steps)-sum(total.steps$steps)
print(paste("Difference in the sum of steps:", sum.difference))
print(paste("% impact:",sum.difference/dim(activity)[1]))
```

### Are there differences in activity patterns between weekdays and weekends?
#### Creating a new factor variable in the dataset with two levels – “weekday” and “weekend”
```{r}
# In Spanish: domingo=sunday, sábado=saturday
week.factor  <- weekdays(as.Date(activityNew$date)
                         , abbreviate=T)=='dom' | weekdays(as.Date(activityNew$date), abbreviate=T)=='sáb'
week.factor  <- factor(week.factor, labels=c("weekday","weekend"))

activityNew  <- cbind(activityNew, week.factor)
head(activityNew)
```

#### Averaged across all weekday days or weekend days for the steps variable
```{r}
average.steps.weekday  <- aggregate(data=activityNew[week.factor=="weekday",], steps ~ interval, FUN="mean")
average.steps.weekend  <- aggregate(data=activityNew[week.factor=="weekend",], steps ~ interval, FUN="mean")
```

#### Panel plot containing a time series plot comparing average of steps 
```{r}
par(mfrow=c(2,1))
plot(x=average.steps.weekday$interval, y=average.steps.weekday$steps, type="l",
     , ylab="Steps", xlab="Interval", main="Average of steps. Weekdays")
plot(x=average.steps.weekend$interval, y=average.steps.weekend$steps, type="l",
     , ylab="Steps", xlab="Interval", main="Average of steps. Weekends")
```


